---
title: "Performance assessment of matches"
author:
  - Barbara Hernando, bhernando@cnio.es
  - Blas Chaves, bchavesu@cnio.es
  - Geoff Macintyre, gmacintyre@cnio.es
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{DemonstratingUtility}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE,comment = "#>")
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)

## PACKAGES
library(dplyr)
library(CNpare)

# DATA DIRECTORY
path_to_data<-paste0(dirname(rstudioapi::getSourceEditorContext()$path),"/data/")
```

In this script, we tested how optimal the matches found by the tool were. 
To this, we applied CNpare to identify the next best match for each of the 604 CCLE cell lines, and then quantified the frequency of significant matches (empirical p-value < 0.05). 
We performed this analysis using both a non-normalized (Manhattan distance) and a normalized (Pearson’s r) similarity measure.

#### 1) Input data

We first get the **RData** files in the `data/` folder:

-   `cells_segcn.RData` -- Absolute copy-number profiles of cancer cell lines generated with *ASCAT*. The segment tables should have the following column headers: "chromosome", "start", "end", "segVal", "sample".

```{r input, include=FALSE}
cells_segcn<-CNpare::cells_segcn
```

#### 2) Get bin tables for CCLE cell lines

Segmented copy-number profiles of cell lines may be fitted to the genomic positions of bins. Therefore, we split genome in evenly sized bins (default 500 kb) and get their positions. 
```{r pos_bins}
allchr=c(1:22) #Add 23 if want to include chrX
lengthChr<-lengthChr
posBins <- lapply(allchr,function(chr) 
    getBinsStartsEnds(window=500000, chr, lengthChr[chr]))
```

We selected the CCLE models with available copy-number profile. We also selected profiles with detectable CIN (those with at least 20 CNAs in a profile). Therefore, we discarded profiles with non detectable chromosomal instability. Finally, a matrix with copy-numbers per bin is then generated.

```{r cnbins}
ccle<-as.data.frame(cbind(cellid=cells_mapping$cellid[cells_mapping$study=="CCLE"],sample=cells_mapping$fileid[cells_mapping$study=="CCLE"]))
cells_segcn <- cells_segcn[cells_segcn$sample%in%ccle$sample,]
cells_segcn <- getCINProfiles(segcn=cells_segcn, samples=unique(cells_segcn$sample))
cells_segcn <- left_join(cells_segcn,ccle, by="sample")[,c(1:4,6)]
colnames(cells_segcn)[5]<-"sample"

#get copy-number data per bin
ccle_cn <- getCNbins(posBins=posBins, data=cells_segcn, samples=unique(cells_segcn$sample))
```

#### 3) Performance assessment of mathces

##### 3.1. Pearson's matches
```{r}
#Get similarities
ccle.measures<-getSimilarities(dat1=ccle_cn, dat2=ccle_cn, method="pearson", pvalue=TRUE)

#Generate a list for the comparison for each case
pearson.list<-list()
for (i in unique(ccle.measures$fileid)){
  element.list<-ccle.measures[which(ccle.measures$fileid==i),]
  pearson.list[[i]]<-as.data.frame(element.list)
  pearson.list[[i]]<-pearson.list[[i]][order(pearson.list[[i]]$r,decreasing = TRUE),] #Order from higher to lower Pearson'r value
  pearson.list[[i]]$sig.pval<-pearson.list[[i]]$r.pval<0.05 #TRUE significative and FALSE non-significative
}

#Determine which of the next best cell line match is significative e.pval<0.05
best.match<-c()
for(i in 1:length(pearson.list)){
  best.match.row<-pearson.list[[i]][2,]
  best.match<-rbind(best.match,best.match.row)
}

#Determine the percentage of significant next best cell line match
percentage.significant<-(nrow(best.match[best.match$sig.pval==TRUE,])/nrow(best.match))*100
print(paste0("% of significant matches: ",round(percentage.significant,2)," (",nrow(best.match[best.match$sig.pval==TRUE,])," cells)"))
```

##### 3.1. Manhattan's matches
```{r}
#Get similarities
ccle.measures<-getSimilarities(dat1=ccle_cn, dat2=ccle_cn, method="manhattan", pvalue=TRUE)

#Generate a list for the comparison for each case
manhattan.list<-list()
for (i in unique(ccle.measures$fileid)){
  element.list<-ccle.measures[which(ccle.measures$fileid==i),]
  manhattan.list[[i]]<-as.data.frame(element.list)
  manhattan.list[[i]]<-manhattan.list[[i]][order(manhattan.list[[i]]$manhattan,decreasing=FALSE),] #Order from lower to higher manhattan distance
  manhattan.list[[i]]$sig.pval<-manhattan.list[[i]]$m.pval<0.05 #TRUE significative and FALSE non-significative
}

#Determine which of the next best cell line match is significative e.pval<0.05
best.match<-c()
for(i in 1:length(manhattan.list)){
  best.match.row<-manhattan.list[[i]][2,]
  best.match<-rbind(best.match,best.match.row)
}

#Determine the percentage of significant next best cell line match
percentage.significant<-(nrow(best.match[best.match$sig.pval==TRUE,])/nrow(best.match))*100
print(paste0("% of significant matches: ",round(percentage.significant,2)," (",nrow(best.match[best.match$sig.pval==TRUE,])," cells)"))
```
